Certifying, incremental type checking

### 2012-06-11 doc/papers/LFMTP2012/paper.tex
paper

\begin{abstract}
 %% 4 sentences:
 %% State the problem
  Type checkers are traditionally implemented as batch processes,
  taking a program and returning a binary information: well-typed or
  not.
 %% Say why it’s an interesting problem
  This laconicism presents a double problem: first, how can I trust
  that the checking algorithm (a complex program) is sound \wrt\ the
  declarative type system? Secondly, how can I avoid costly rechecks
  of the whole program when I only make a small, localized change?
 %% Say what your solution achieves
  A lightweight solution to both problems is letting the checker
  return a typing derivation: trust is then ensured by checking that
  this derivation is well-formed, and incrementality by the
  possibility of reusing already-computed pieces of derivations.
 %% Say what follows from your solution
  We present a system turning a type system and its checking algorithm
  into a safe and incremental one, consisting of a generic data
  structure to represent reusable derivations based on \LF\ and a
  general-purpose programming language to generate them.
\end{abstract}

@@

\section{Use case}
%% It’s an unsolved problem (1 page)

\subsection{A backend for incremental type checking}
\label{sec:use-incremental}

\begin{figure}
  \centering
%  \small
  \begin{mathpar}
    \infer{ }{\jts{\cst o} {\even}}

    \infer{\jts M \odd} {\jts {\s M} \even}

    \infer{\jts M \even} {\jts {\s M} \odd}

    \infer{\jts M \nat} {\jts {\s M} \nat}

    \infer{\jts M {A\to B} \and \jts N A} {\jts {\app M N} B}

    \infer{\infer*{}{\mbox{$
          \begin{array}{c}
            [\jts {\var x} A] \\[-0.3em]
            \vdots\\
            \jts M B
          \end{array}$
        }}} {\jts {\lam x M} {A\to B}}

    \infer{\jts M A \and \jsub A B}{\jts M B}

    \infer{\jts M \nat
      \and
      \jts N A
      \and
      \infer*{}{\mbox{$
        \begin{array}{c}
          [\jts {\var x} \nat] \quad
          [\jts {\var y} A] \\[-0.3em]
          \vdots \\
          \jts P A
        \end{array}
        $}}
    }{\jts {\recb M N x y P} A}

    \infer{ }{\jsub A A}

    \infer{ }{\jsub \even\nat}

    \infer{ }{\jsub \odd\nat}

    \infer{\jsub {A'} A \and \jsub B {B'} }{\jsub {A\to B}{A'\to B'}}
  \end{mathpar}
  \caption{System \sysname{T_{<:}}, declaratively}
  \label{fig:tsub}
\end{figure}

A student is learning about System \sysname {T_{<:}} --- a simply
typed $\lambda$-calculus with natural numbers $\cst{nat}$ coming in
two sets $\cst{even}$ and $\cst{odd}$, introduction $\z$ and $\s{M}$,
elimination ($\recb M N x y P$), and subtyping on those types; see
Fig. \ref{fig:tsub}. He (or she) wants to write a program computing
$(2+2)\times 3$. He knows that sometimes those $\cst{rec}$ arguments
are quite hard to get correct, but \sysname{T_{<:}} is a typed
language; he wants to take advantage of this to build his program
incrementally using the teacher's provided incremental verifier. As a
first attempt, he wants to verify $P_1$:
$$
\recb{\s\z}{\s\z} x y {\s {\var x}}\ \text{.}
$$
For this, he sends the command $\finfer{P_1}$ to the system. The
$\finfer{}$ operator takes a well-formed program and tries to construct
a valid typing derivation for its argument in the current, empty
context. The system succeeds and returns a typing derivation $\md_1$
of \/ $\vdash P_1 : \cst{nat}$, valid in an empty context.

@@

\subsection{A language to write incremental type-checkers}

@@

\section{Here is my idea}
%% Here is my idea (2 pages)

@@

The emission of certificates by untrusted applications first raises
the question of the \emph{data structure} to use for these
certificates. The requirements are:
\begin{itemize}
\item conciseness of the certificates, since they are stored or even
  transferred remotely;
\item a well-understood, short and independent verification algorithm.
  This is known as the \emph{de Bruijn principle};
\item universality with respect to the particular problem: having a
  certificate language for each domain of application would just shift
  the problem of trust; having one universally accepted certificate
  language that can be tailored to many problems puts a much stronger
  trust on the unique verifier.
\end{itemize}

The Edinburgh Logical Framework \cite{harper1993framework}, or \LF\
for short, is a well-accepted solution to this question. It is a typed
lambda-calculus and can be viewed as a language for
\emph{representing} terms, derivations and statements of $\Pi_0^1$
meta-theorems (but not their proofs) in a given inference system
supporting hypothetical judgments \cite{pfenning2001logical}. For
these, it provides built-in support for encoding languages with
binders and hypothesis (a feature called $\lambda$-tree syntax or HOAS
\cite{pfenning1988higher}). The user gives a \emph{signature} defining
abstract syntax and inference rules of his \emph{object language}, and
gets an algorithm to verify terms and derivations expressed in the
\LF\ syntax.

For instance, the rule for System \sysname{T}'s recursor can be
encoded by the following constant:

\begin{mathleft}
  \cst{is\_rec}\ :\
  \prd{M}{\cst{tm}}\prd{N}{\cst{tm}}\prd{P}{\cst{tm}\to\cst{tm}\to\cst{tm}}
  \\\quad
  \prd A {\cst{tp}}
  \app {\cst{is}}\app M {\cst{nat}} \to \app{\cst{is}}\app N A \to \\\quad
  \left(
  \prd x {\cst{tm}} \prd y {\cst{tm}}
  \app{\cst{is}}\app x \cst{nat} \to \app{\cst{is}}\app y A \to
  \app{\cst{is}}\app {(\app P \app x y)} A
  \right)\to\\\quad
  \app{\cst{is}}\app{(\app{\cst{rec}}\app M \app N (\lam x\lam y \app
    P\app x y))} A
\end{mathleft}

This framework has been used to issue certificates for a
number of practical applications like proof-carrying code
\cite{necula1997proof}, authentication framework \cite{appel1999proof}
\ldots %TODO ATP

To write deltas by sharing subterms, we need to be able to address any
subterm and instantiate its free variables. For this, we define the
\emph{slicing} of a well-typed \LF\ term: we extend \LF\ with
contextual metavariables $\meta X$, $\meta Y$\ldots\ as in
\cite{nanevski2008contextual}. Metavariables stand for well-typed
terms defined in a repository, and are instantiated wrt. their free
variables. A repository also contains a distinguished metavariable
corresponding to the ``tip'' of the term stored in it. For instance, the
repository
$$
\small
\left(\begin{array}{llll}
    \meta X &\mapsto
    \envnil&\vdash
    \app{\cst{lam}} \lam x \smeta Y {\msubst u {\smeta T {\msubst w x}}} &:
    \cst{tm}
    \\
    \meta Y &\mapsto
    \env u {\cst{tm}} &\vdash
    \app{\cst{lam}} \lam y \smeta Z {\msubstcons {\msubst u
      {\var u}} v {\var y} } &:
    \cst{tm}
    \\
    \meta Z &\mapsto
    \envcons {\env u {\cst{tm}}} v {\cst{tm}} &\vdash
    \app{\app{\cst{app}}{\var v}}{\var u} &:
    \cst{tm}
    \\
    \meta T &\mapsto
    \env x {\cst{tm}} &\vdash
    \app{\cst{app}}\app{\var x}{\cst{o}} &:
    \cst{tm}

\end{array}\right), \smeta X {}
$$
is one of the possible way of slicing of the term
$$
\fbox{$
  \app {\cst{lam}} \lam x
  \fbox{$
    \app {\cst{lam}} \lam y
    \fbox{$
      \app {\cst{app}} \app {\var y}
      (\fbox{$
        \app{\cst{app}}\app{\var x}{\cst o}
      $})
    $}
  $}
$}
$$
where all boxed subterms are referable slices.

@@

@@

The value calculus defines the \emph{values} we want to
manipulate. These contain constants, and --- contrarily to \eg\
\sysname{ML} values --- binders and variables enabling $\lambda$-tree
syntax. Thus, they embed a superficial notion of computation:
hereditary substitution. This calculus is an extension of \emph{spine
  canonical \LF} \cite{pfenning2007term} enriched with metavariables
which are references to open terms as in
\cite{nanevski2008contextual}. We present its typing algorithm
relatively to a signature. It features an explicit caching mechanism
using metavariables.

@@

    \infer[UEval]{
      \Sigma(c) = A; \anndefined T
      \and
      \jcc\Delta\Gamma T S A {\Delta'} F
    }{
      \jcr\Delta\Gamma{\app {\cst c} S} \Delta {\app {\cst c} S}
    }

>>>>

\begin{abstract}
 %% 4 sentences:
 %% State the problem

  A lightweight way to design a trusted type-checker is to let it
  return a \emph{certificate} of well-typing (for instance
  \sysname{Agda} and \sysname{GHC} adopt this architecture
  internally).  Most of the time, these type-checkers are confronted
  sequentially with very similar pieces of input: the same program,
  each time slightly modified. Yet they rebuild the entire
  certificates, even for unchanged parts of the code. Instead of
  discarding them between runs, we can take advantage of these
  certificates to improve the reactivity of type-checking, by feeding
  parts of the certificates for unchanged parts of the original
  program back into the new input, thus building an incremental
  type-checking.

  We present a language-independent framework making possible to turn
  a type system and its certifying checking algorithm into a safe,
  incremental one. It consists in a model of incrementality by
  sharing, a generic data structure to represent reusable derivations
  based on contextual \LF, and a general-purpose language to write
  certifying programs.
\end{abstract}

@@

\section{Using the framework}
%% It’s an unsolved problem (1 page)

\subsection{As a programmer}
\label{sec:use-incremental}

% \begin{figure}
%   \centering
% %  \small
%   \begin{mathpar}
%     \infer{ }{\jts{\cst o} {\even}}

%     \infer{\jts M \odd} {\jts {\s M} \even}

%     \infer{\jts M \even} {\jts {\s M} \odd}

%     \infer{\jts M \nat} {\jts {\s M} \nat}

%     \infer{\jts M {A\to B} \and \jts N A} {\jts {\app M N} B}

%     \infer{\infer*{}{\mbox{$
%           \begin{array}{c}
%             [\jts {\var x} A] \\[-0.3em]
%             \vdots\\
%             \jts M B
%           \end{array}$
%         }}} {\jts {\lam x M} {A\to B}}

%     \infer{\jts M A \and \jsub A B}{\jts M B}

%     \infer{\jts M \nat
%       \and
%       \jts N A
%       \and
%       \infer*{}{\mbox{$
%         \begin{array}{c}
%           [\jts {\var x} \nat] \quad
%           [\jts {\var y} A] \\[-0.3em]
%           \vdots \\
%           \jts P A
%         \end{array}
%         $}}
%     }{\jts {\recb M N x y P} A}

%     \infer{ }{\jsub A A}

%     \infer{ }{\jsub \even\nat}

%     \infer{ }{\jsub \odd\nat}

%     \infer{\jsub {A'} A \and \jsub B {B'} }{\jsub {A\to B}{A'\to B'}}
%   \end{mathpar}
%   \caption{System \sysname{T_{<:}}, declaratively}
%   \label{fig:tsub}
% \end{figure}

A student is learning about System \sysname {T_{<:}} --- a simply
typed $\lambda$-calculus with natural numbers $\cst{nat}$ coming in
two sets $\cst{even}$ and $\cst{odd}$, introduction $\z$ and $\s{M}$,
elimination ($\recb M N x y P$), and subtyping on those types. He (or
she) wants to write a program computing $(2+2)\times 3$. He knows that
sometimes those $\cst{rec}$ arguments are quite hard to get correct,
but \sysname{T_{<:}} is a typed language; he wants to take advantage
of this to build his program incrementally using the teacher's
provided incremental verifier. As a first attempt, he wants to verify
$P_1$:
$$
\recb{\s\z}{\s\z} x y {\s {\var x}}\ \text{.}
$$
For this, he sends the command $\finfer{P_1}$ to the system. The
$\finfer{}$ operator takes a well-formed program and tries to construct
a valid typing derivation for its argument in the current, empty
context. The system succeeds and returns a typing derivation $\md_1$
of \/ $\vdash P_1 : \cst{nat}$, valid in an empty context.

@@

\subsection{As a type system designer}
\label{sec:lang-incr}

@@

\section{Designing the framework}
%% Here is my idea (2 pages)

@@

The emission of certificates by untrusted applications first raises
the question of the \emph{data structure} to use for these
certificates. % The requirements are:
% \begin{itemize}
% \item conciseness of the certificates, since they are stored or even
%   transferred remotely;
% \item a well-understood, short and independent verification algorithm.
%   This is known as the \emph{de Bruijn principle};
% \item universality with respect to the particular problem: having a
%   certificate language for each domain of application would just shift
%   the problem of trust; having one universally accepted certificate
%   language that can be tailored to many problems puts a much stronger
%   trust on the unique verifier.
% \end{itemize}
The Edinburgh Logical Framework \cite{harper1993framework}, or \LF\
for short, is one of the well-accepted solution to this question. It
is a typed lambda-calculus and can be viewed as a language for
\emph{representing} terms, derivations and statements of $\Pi_0^1$
meta-theorems (but not their proofs) in a given inference system
supporting hypothetical judgments \cite{pfenning2001logical}. For
these, it provides built-in support for encoding languages with
binders and hypothesis (a feature called $\lambda$-tree syntax or HOAS
\cite{pfenning1988higher}). The user gives a \emph{signature} defining
abstract syntax and inference rules of his \emph{object language}, and
gets an algorithm to verify terms and derivations expressed in the
\LF\ syntax.
% For instance, the rule for System \sysname{T}'s recursor can be
% encoded by the following constant:
% \begin{mathleft}
%   \cst{is\_rec}\ :\
%   \prd{M}{\cst{tm}}\prd{N}{\cst{tm}}\prd{P}{\cst{tm}\to\cst{tm}\to\cst{tm}}
%   \\\quad
%   \prd A {\cst{tp}}
%   \app {\cst{is}}\app M {\cst{nat}} \to \app{\cst{is}}\app N A \to \\\quad
%   \left(
%   \prd x {\cst{tm}} \prd y {\cst{tm}}
%   \app{\cst{is}}\app x \cst{nat} \to \app{\cst{is}}\app y A \to
%   \app{\cst{is}}\app {(\app P \app x y)} A
%   \right)\to\\\quad
%   \app{\cst{is}}\app{(\app{\cst{rec}}\app M \app N (\lam x\lam y \app
%     P\app x y))} A
% \end{mathleft}
This framework has been used to issue certificates for a number of
practical applications like proof-carrying
code~\cite{necula1997proof}, authentication
framework~\cite{appel1999proof} \ldots %TODO ATP

To write deltas by sharing subterms, we need to be able to address any
subterm. Since a subterm can be open, we must instantiate its free
variables. For this, we define the \emph{slicing} of a well-typed \LF\
term: we extend \LF\ with contextual metavariables $\meta X$, $\meta
Y$, \ldots\ as in \cite{nanevski2008contextual}. Metavariables stand
for well-typed terms defined in a repository, and are instantiated
wrt. their free variables. A repository also contains a distinguished
metavariable corresponding to the ``tip'' of the term stored in
it. For instance, the repository
$$
\small
\left(\begin{array}{llll}
    \meta X &\mapsto
    \envnil&\vdash
    \app{\cst{lam}} \lam x \smeta Y {\msubst u {\smeta T {\msubst w x}}} &:
    \cst{tm}
    \\
    \meta Y &\mapsto
    \env u {\cst{tm}} &\vdash
    \app{\cst{lam}} \lam y \smeta Z {\msubstcons {\msubst u
      {\var u}} v {\var y} } &:
    \cst{tm}
    \\
    \meta Z &\mapsto
    \envcons {\env u {\cst{tm}}} v {\cst{tm}} &\vdash
    \app{\app{\cst{app}}{\var v}}{\var u} &:
    \cst{tm}
    \\
    \meta T &\mapsto
    \env x {\cst{tm}} &\vdash
    \app{\cst{app}}\app{\var x}{\cst{o}} &:
    \cst{tm}

\end{array}\right), \smeta X {}
$$
is one of the possible way of slicing the term
$$
\fbox{$
  \app {\cst{lam}} \lam x
  \fbox{$
    \app {\cst{lam}} \lam y
    \fbox{$
      \app {\cst{app}} \app {\var y}
      (\fbox{$
        \app{\cst{app}}\app{\var x}{\cst o}
      $})
    $}
  $}
$}
$$
where all boxed subterms are referable slices.

@@

Secondly, we need a mean to transform our \LF\ objects as in
\ref{sec:lang-incr}. For this we introduce \emph{interpreted
  functions} $\interp f$, $\interp g$, \ldots in the signatures,
besides constants. Each $\interp f$ comes with its type and its code:

@@

The value calculus defines the \emph{values} we want to
manipulate. These contain constants, and --- contrarily to \eg\
\sysname{ML} values --- binders and variables enabling $\lambda$-tree
syntax. Thus, they embed a superficial notion of computation:
hereditary substitution. This calculus is an extension of \emph{spine
  canonical \LF} \cite{pfenning2007term} enriched with metavariables
which are references to open terms as in
\cite{nanevski2008contextual}. We present its typing algorithm
relatively to a signature. % It features an explicit caching mechanism
% using metavariables.

@@

    \infer[UEval]{
      \Sigma(c) = A; \anndefined T
      \and
      \jcc\Delta\Gamma T S A {\Delta'} F
    }{
      \jcr\Delta\Gamma{\app {\cst c} S} {\Delta'} F
    }
